---
title: 'Lab 4'
subtitle: "Due: 10/24/2019 by 10 pm"
author: "Aashwin Lamsal, Eric Friesth"

date: "10/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Some abbreviations for the column names can be found in the Glossaries sheet in the Excel file.

**References:**

- [Iowa State University Athelete](https://cyclones.com/index.aspx)
- [A Beginner's Guide to American Football](https://www.youtube.com/watch?v=3t6hM5tRlfA) for the rules of American football.

## Loading Data: Cyclone Football 
```{r}
library(readxl)
defense <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
str(defense)
```

```{r,echo=FALSE,results='hide'}
offense <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')
bio <- read_excel('cyclonesFootball2018.xlsx', sheet='Biography')
```

## Part one: Cleaning data
The three Excel sheets are now loaded into R but they are not useful to us before some data cleaning. So we must manipulate the three imported data frames. The manipulation of the data must fit the following requirements

  -The player names and opponent names are factors.
  -All offensive and defensive statistics are numerical. The `Weight` column in Biography is also numeric.  
  -Separate the `Height` column into two columns. The newly created columns must be numeric. Then we create one column that     
   represents `Height`, dropping the irrelevant columns in the process.

  we are then left with three clean data frames: `defClean`, `offClean`, and `bioClean` with the structures of each being printed below.


```{r}
library(tidyverse)
library(dplyr)
#Part1: player names and opponent names are factors
offense$Name <- factor(offense$Name)
defense$Name <- factor(defense$Name)

offense$Opponent_Opponent <- factor(offense$Opponent_Opponent)
defense$Opponent_Opponent <- factor(defense$Opponent_Opponent)

str(offense) #checking to see if the changes were made successfully
str(defense)

#Part 2: All offensive and defensive statistics are numerical. The `Weight` column in Biography is also numeric. (Hint: look at `?mutate_at` if you want to save writing some code).

offClean<- offense %>% mutate_at(c("Receiving_REC","Receiving_YDS","Receiving_TD","Rushing_ATT","Rushing_YDS", "Rushing_TD", "Passing_CMP-ATT", "Passing_YDS", "Passing_TD", "Passing_INT"), funs(as.numeric), na.rm=TRUE)

str(offClean) #checking if changes were made successfully

defClean<- defense %>% mutate_at(c("Tackles_Solo", "Tackles_ASST", "Tackles_TFL","Tackles_Sack","Turnover_FF", "Turnover_FR", "Turnover_INT", "Pass_QBH", "Pass_PB"), funs(as.numeric), na.rm=TRUE)

str(defClean)#checking if changes were made successfully

bio <- bio%>% mutate_at(c("Weight"), funs(as.numeric))
str(bio)

#Part 3: Separate the `Height` column into two columns. Make sure the newly created columns are numeric. Then create one column that represents `Height` (Hint: Use a unit that works well). Drop any irrelavant columns from this process

newBio = bio

bio_split = strsplit(as.character(newBio$Height), split = "-")
newBio = transform(newBio, Height_Feet = sapply(bio_split, "[[", 1), Height_Inches = sapply(bio_split, "[[", 2))
str(newBio)
#newBio = newBio %>% mutate_at(c("Height_Feet", "Height_Inches"), funs(as.numeric))

newBio= newBio%>% mutate(heightInFeet= as.numeric(as.character(newBio$Height_Feet))) #since splitting the height into 2 columns gives us two columns of factors, converting to character then numeric is a workaround for the type conversion error caused by directly converting from factor type
newBio= newBio%>% mutate(inchesInChar=as.numeric(as.character(newBio$Height_Inches)))

newBio= newBio%>% mutate(totalHeighInInches= (heightInFeet * 12) + inchesInChar) #new calculated column for total heights

newBio <- newBio%>% select(-c("Height_Feet","Height_Inches", "Highschool", "Height", "heightInFeet", "inchesInChar"))

str(newBio)


bioClean = newBio


#Part 4: By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: `defClean`, `offClean`, and `bioClean`. Print the structures of these data frames. **Make sure your data types make sense for each variable.**

str(bioClean)
str(defClean)
str(offClean)

```



## Part two: Tidying 


Next, to tidy the tables, we must reformat the `defClean` data frame into a tidy format by using `gather`. After this, we created a histogram for each defensive statistic, and used `facet_wrap` to draw all histograms in a single plot. Based on the distributions of the defensive skills below, we see that many of these skills are rare such as turnovers (specifically forced fumbles, fumble recovery, and interceptions) as well as tackles (specifically sacks).

We then moved onto the question of whether or not ISU had a better defense against West Virginia or Kansas State. We answered this question by creating a scatterplot with x-axis being the name of each player, the y-axis the number of solo tackles each player made, and our color coding indicating which points represent the Kansas State and West Virginia games respectively (please excuse us for this is not exactly what the question was asking for, we ended up getting really stuck here to be quite honest). A large number of solo tackles indicates good defense, and juding from the scatter plot we made below, we observed the number of solo tackles made by each player on the Iowa State Football active roster. This means that even offensive players who weren't going to be going for any defensive plays, let alone solo tackles if the offense is playing well, are going to be shown upon this plot. However, from looking at the players who did make key solo tackles, we saw that there were more solo tackles made in the game against Kansas State than agains West Virginia, especially after looking and seeing that the highest amount of solo tackles against West Virginia was 5, whereas 3 players made 5 or more solo tackles in the Kansas State game. We conclude that Iowa State's defense played better against Kansas State, due to the impact of the number of solo tackles made.


```{r}

defClean = defClean %>% gather(key = Statistic, value = score, 3:11)


ggplot(defClean, aes(x = defClean$score)) + geom_histogram(bins = 350, binwidth = 1) + facet_wrap(~Statistic) + labs(title = "Distribution of Defensive Skill Statistics by their Score", x = "Score (10 point scale)", y = "Frequency of Defensive Skill")


dfWVKState <- defClean %>% filter((defClean$Opponent_Opponent == "West Virginia" & defClean$Statistic == "Tackles_Solo") |(defClean$Opponent_Opponent == "Kansas State" & defClean$Statistic == "Tackles_Solo" ))


#newDf <- data.frame(defKState$score)
#newDf <- newDf%>% mutate(KWVeTackles=defWV$score)

ggplot(dfWVKState, aes(x=Name, y=score, color = Opponent_Opponent)) + geom_point() + theme(axis.text.x =element_text(angle = 90, hjust= 1)) + labs(title = "Spread of Solo Tackles against Kansas State and West Virginia by each player", y = "Solo Tackles Made")+scale_fill_discrete(name = "Opponent")

```

Now we turn our attention to the player biography `bioClean`. First, we separate the `Hometown` column into two columns, namely the city and the state, and print the first few rows of these two columns. We separated the hometown by city and state, through using the same technique as before, all in part to the strsplit function, then using transform to create a new data frame that includes the city and state columns we created. We then grouped the players by their home state, and counted how many of the players hail from each state (see below).

```{r}
hometownSplit = strsplit(as.character(bioClean$Hometown), split = ", ")
newBioClean = transform(bioClean, city = sapply(hometownSplit, "[[", 1), state = sapply(hometownSplit, "[[", 2))
head(newBioClean)
newBioClean %>% group_by(newBioClean$state) %>% count()

```


## Part three: Joining data frames

The code below joins the offense and biography tables at an outer-left-join, and viewing the head will display hometowns, as well as cities and states individually.


```{r}
#joining the offensive player data together with the player biography 
library(dplyr)
jointOffense<- left_join(offClean, newBioClean, by= "Name") #left joined the player biography onto the offense table

jointOffense<- jointOffense%>% group_by(Name)
head(jointOffense, 50)


```

Regarding the relationship between receiving yards and an offensive player's weight, as you can see from the bar chart below, wide receivers (typically of weight 225lbs) have more receiving yards than any other position or weight. There is a positive correlation between being a wide receiver and being 225lbs as well as a positive correlation between being a wide receiver and having more receiving yards.

```{r}

#ggplot(jointOffense, aes(x = jointOffense$Weight, y = jointOffense$Receiving_YDS, color = jointOffense$Position)) + geom_point(position = "jitter") + labs(title = "Relationship between a Player's Weight and their Receiving Yards", x = "Player Weight (lbs)", y = "Player Receiving Yards")

ggplot(jointOffense, aes(x = jointOffense$Weight, y = jointOffense$Receiving_YDS, fill = jointOffense$Position, width = 2)) + geom_bar(stat = "identity") + labs(title = "Relationship between a Player's Weight and their Receiving Yards", x = "Player Weight (lbs)", y = "Player Receiving Yards")

```
 Going off of the previous question, my lab partner and I decided it would be interesting to find the demographics of each player position by looking for clusters (through a scatter plot) based on height and weight of the player. Based on your height and weight, what position are you most likely to be in? So we made a plot and from what we can see on our scatter below, there are very distinct clusters for Height-to-Weight ratios that you would probably want to fit into in order to play that offensive position.

```{r}

ggplot(jointOffense, aes(x = jointOffense$Weight,  y = jointOffense$totalHeighInInches, color = jointOffense$Position)) + geom_point(position = "jitter") + labs(title = "Weight to Height Correlation for Offensive Postions", x = "Weight (pounds)", y = "Height (inches)")



```
